Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-07-09 MS/ms.Rmd
Finished off text. Corrected spelling. Need to add survey and do supplement and first draft should be ready

A meta-analysis provides quantitatively informed answers to research questions by synthesising research results across studies and identifying sources of variation [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Arnqvist1995; @Koricheva2013; @Borenstein2019]. Meta-analysing independent studies provides greater statistical power and precision than what any individual study on its own would be able to provide on a question - particularly given that most empirical studies are already under-powered in many areas of biology [@Forstmeier2017; @Jennions2003; @Button2013]. By expressing study effects on a common scale (i.e., standardised effect size) a meta-analyst can gain broader insight into the direction and efficacy of a particular treatment effect or the strength of a relationship between two variables of interest [@Gurevitch2018; @Koricheva2013]. Quantitative syntheses, including meta-analyses, have provided comparative physiologists with powerful insights on pressing global challenges - from testing whether physiological plasticity can buffer organisms against climate change [e.g, @Seebacher2015] to the degree to which endocrine disrupting chemicals, such as bisphenol A (BPA), impact aquatic organisms [e.g., @WuSeebacher2020].

@@

Despite its widespread adoption and well established methodological procedures, meta-analysis is often criticised for mixing 'apples and oranges' -- in other words, mixing effects from studies that lack comparability [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010; @Glass2015]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units, which can impact the generalisability of meta-analytic results [@Arnqvist1995; @Stewart2010]. Indeed, study specific differences in comparative physiology are widespread. For example, experimental designs can vary in the temperatures that they apply or in the dosages of chemical modifiers used in experimental treatments [e.g., @WuSeebacher2020]. To the uninitiated, this concern might not be unfounded given that heterogenity in effects is often high in ecological and evolutionary meta-analyses [@Senior2016], telling us that the effect size varies a great deal from one population to another. However, to many, the very fact that highly heterogeneous effects exist is a core strength of meta-analysis [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015]. Indeed, exploring the reasons for why effects are likely to vary across studies "...is the spice of life" [p.g., 519; @Cooperetal2019].

The goal of our review is to briefly overview the process of meta-analysis, making a case for why heterogenity in a meta-analysis is a good thing for answering fundamental reserch problems and guiding research directions. We introduce the idea of 'nuisance heterogenity', showing comparative physiologists that it can be comfortably dealt with by re-formalising many existing effect sizes measures and/or by using multilevel meta-regression models. Indeed, a combination of these approaches might often be ideal. We first conduct a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic practices -- determining what types of effect sizes are commonly used, how often effect size estimates are impacted by 'nuisance heterogeneity' and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. We then formalise alternative effect sizes and their associated sampling variance to provide comparative physiologists opportunities to explicitly incorporate treatment differences into the effect size estimates to ease their interpretation. We then describe how more complex treatment differences, such as dosage differences (which may have non-linear effects) can be accommodated using multilevel meta-regression models. We hope to show how expanding the meta-analytic toolkit for comparative physiologists will provide new opportunities to address global scale problems that attempt to understand how organisms will cope with rapidly changing environments in the future.

@@

Generally, contributors to effect size heterogeneity fall within a number of broad categories. First, sampling variance contributes to heterogeneity; the variability driven by the inherent uncertainty in sampling a given population. We cannot sample the entire population and nor do we necessarily want to. As a consequence, we need to rely on a sub-sample to estimate a population parameter (e.g., correlation between two traits). How many independent samples we collect impacts our precision in estimating this 'true' effect [@Borenstein2019]. In meta-analysis we formally include sampling variance to remove this source of heterogeneity directly -- hence why having sampling variances for an effect is so important [@Gurevitch2018; @Koricheva2013]. Second, there are sociological factors impacting heterogeneity. These include attitudes towards publishing non-significant results. Low powered studies that do not find an 'effect' or that find an effect in the opposite direction often lead to what is called publication bias [@Jennions2013; @Nakagawa2021b; @Rothstein2005]. Third, there is real biology driving differences in effects. This is particularly true in comparative physiology where different species with varying habitats, life-histories, mating systems and reproductive modes are synthesised. As comparative physiologists, this is the stuff that gets us up every morning! Much variation could result from shared evolutionary history (i.e., phylogeny) or because of the different thermal environments species (or populations) occupy. We can model these biological factors to test predictions from hypotheses that we believe are key players impacting mean effect size and direction. This improves our understanding of the biological world around us, and helps inform on future research directions. Finally, methodological factors are also big contributors to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it is important we know and understand this as it can determine how experiments are designed in the future. In other cases, methodological factors might be considered 'nuisance heterogeneity' [following from @Cooperetal2019] -- factors we know vary, but are rather unsurprising to us. These can include differences in temperature, pH, dosages, water potential etc. The factors are obviously going to vary and have an impact on the magnitude of an effect, but they might distract from the core interest. In some cases, however, these factors might actually not be 'nuisance' variables but rather the main interest of a synthesis. However, in many cases they can result in a more complicated interpretation of an effect measure then what is desired.

All the various sources of heterogeneity (except sampling variance) all contribute (to varying extents) to the 'apples and oranges' problem often touted as being an issues with some meta-analyses (or quantitative syntheses more generally). Indeed, some of this criticism revolves around combining studies that vary in their quality (e.g., observational vs experimental). Combining standardised effects, while ignoring these sources of heterogeneity can result in the meaning of a meta-analysis not being particularly useful to many. However, many argue that this is just what we are interested in synthesizing [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015]. In the words of Gene Glass, the father of meta-analysis: "Of course it mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. @Glass2015's intention was to simply highlight that whether comparing apples and oranges is a good idea simply boils down to the specific question of interest, the population one wishes to make conclusions about and how hetergenous effects actually are. If we are interested in gaining insight into 'fruit' then synthesising apples and oranges is the only thing that makes sense. Instead, if we are interested in generalising only to apples then mixing fruit will not be ideal. In addition, if we want to understand why different 'fruit' vary in their effects, then we can only ask this question by including all fruit [@Borenstein2019].

We tend to agree with @Glass2015, and his sentiment is of particular relevance to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms then including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are known to differ. By synthesizing diverse study outcomes, it allows us to not just report on overall effects (if the question is relevant to ask), but to understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction) and approaches that can explain the diversity of effect outcomes observed. In doing so, it provides a richer set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included within a meta-analysis, and how this can affect interpretations made -- an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. If the estimate of an overall effect coincides with low heterogenity estimates then the effects are reasonably consistent [@Borenstein2019]. If, however, an overall effect is accompanied by large heterogenity estimates, then we would shift focus and explore drivers of that heterogeneity. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant, and indeed may be misleading. Therefore, understanding overall effects in each of these groups seperately (and whether they differ) will be more useful. It is important to consider these questions when conducting meta-analyses.

Of course, there are certain types of heterogeneity that transcend different questions or that are simply a 'nuisance' to interpreting effects. In other words, if we compare different types of 'apples' is it still fair to synthesize these studies to understand overall effects? This situation might be analogous to a meta-analysis in comparative physiology that attempts to understand the effects of dietary on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of fish under different temperatures, but in the end, they are still all the same fish species, governed by the same thermodynamic effects on growth. In this situation, we might still be interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. It is not always clear how we should tackle these sources of heterogeneity. If temperature differences are of direct interest, then we should definitely model them. If, however, it is less interesting, than say the diet administered, then it may introduce an unnecessary complication to the interpretation of the effect(s) of interest.

@@

As discussed above, different types of questions and experimental approaches will mix 'apples and oranges' to varying extents, and indeed, this heterogeneity (or basket of fruit) is often the main interest to comparative biologists. To better understand the different types of effect sizes being used, their susceptibility to known sources of methodological / environmental variability across studies, and how these sources are controlled we conducted a literature survey. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to continuous variation in how treatments were applied (e.g., for contrast-based effects) or the conditions under which traits were measured (i.e., for correlations) across studies? 3) If so, was the variation resulting from treatment differences accounted for in the effect size or during analysis?, and finally, 4) how were effect sizes analysed in the quantitative synthesis, were these weighted analyses, and what program was used for analysis? We also recorded whether publication bias and sensitivity analyses were conducted, if data were publicly available, where it was stored, and if so, if it was usable (i.e., in a machine readable format).

@@

All searches were conducted in Scopus between May 16 - 18, 2021. Briefly, we first did a series of pilot searches (detailed in Supplementary Materials) to refine our search query [See @Foo2021 for an overview of this process]. We searched for studies that included: 'meta-analysis', 'meta regression' , 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macro physiology' in their title and abstract, restricting our search to key comparative physiology journals (detailed in Table X - Supplementary Materials) within the last 6 years. We expected this to provide a contemporary overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We acknowledge that this may have missed some important, and influential, quantitative synthesis in the field of comparative physiology, but likely provided a random sample to inform on the questions we sought to address (see below).

We included quantitative syntheses, including meta-analyses (*sensu stricto*), that were focused on comparative physiological questions only. Narrative reviews and comments were not considered because they did not collate quantitative data from the literature. Studies needed to have collected physiological trait data from the literature including traits, such as hormones, thermal physiology (CTmax, CTmin, thermal preferences, thermal breadth, LT50, lethal thresholds), metabolic rates (routine, maximum, and minimum metabolic rate, aerobic scope), enzyme reaction rates and concentrations, oxidative stress and consequences (e.g., DNA damage, teleomere attrition), performance measurements (e.g., heart rates, sprint speed, swim speed, growth rates), body composition (e.g., fatty acid profiles) and immune traits (e.g., cell function, antibodies, ect.). Studies were also included if they measured the relationship between physiological traits and body size (e.g., metabolic scaling), reproductive allocation, or survival / mortality across taxa. Studies on behavior were only included if the synthesis was explicitly interested in the proximate physiological drivers of behavioral variation. We also included meta-analyses that quantified plasticity in physiological traits across taxa for the purpose of understanding adaptive plasticity in response to environmental variation. While many studies in comparative physiology collect data as part of a comparative analysis, we focus here only on studies collating data from existing literature because this is the major data source for quantitative synthesis. Quantitative syntheses interested in phenology, bio-mechanics, sexual size dimorphism (i.e., SSD, except if the study was interested in sexual differences in physiological traits), those focused on validating physiological methodology, or characterising population genetic structure, gene transcription and quantitative genetics were excluded. There were also meta-analyses that focused on colour evolution and contrast which varied in their goals but were most often not focused on physiological questions. As such, these studies were also exclude. We restricted our focus to animal studies because plant synthesis papers were often only focused on ecological (e.g., quantified changes in abundance and distribution of species) or agricultural questions (e.g., studies on soil microbial function, nitrogen and phosphorus capture, methane output and greenhouse gas emissions).

All title, abstract and full text screening was done by two people (DWAN and ML). Any conflicts or disagreements were resolved by discussion and a consensus on a paper was reached. Out of the 417 papers, 79 full texts were screened and we included a total of 62 for final data extraction. For full details on the search strings used, dates searched, the number of papers found, and our PRISMA flowchart (along with reasons for exclusion at the full text stage) we refer readers to the *Supplementary Materials*. To address our questions, we categorised effect sizes as falling into the following categories: Response Ratio (lnRR), Standardised Mean Differences (SMD; e.g., Hedges d, g or Cohen's d), Correlation or Fisher's Z -transformed Correlation, Regression Slopes (Slopes), Risk Ratios, or whether the paper analysed the raw data or summary statistics directly (i.e., means/proportions) (which we label as raw/mean/proportion). For all other effects we labelled as 'other'. To determine whether effect sizes were likely to vary based on some underlying continuous variable or were dependent on the context in which traits were measured, we assessed whether authors: 1) acknowledged this explicitly in the paper or 2) accounted for it in their models. If it was not clear, two of us (DWAN & PP) discussed the paper, and made a decision. For example, trait means, such as ectothermic metabolic rate (i.e., raw/mean/proportion-based effect size), will vary depending on the temperature being measured, the magnitude of differences between two treatments (e.g, contrast-based effect sizes - SMD or lnRR) might depend on the temperature or dosage used in treatments, and the strength of relationship between two variables (e.g., correlation between behaviour and metabolism) might depend on the temperature context under which metabolic rate or behaviour were measured. Given the high heterogeneity in effects used in comparative physiology, we expected this to be common to most effect sizes, however, it has not been formally quantified. We categorised the types of staitsical models fit in each study as: 1) fixed-effect meta-analysis; 2) random effects meta-analysis (i.e., intercept only model, but has random study effects); 3) multi-level meta-analysis (i.e., intercept only model, but has multiple random effect levels such as study, species, observation-level random effects); 4) meta-regression models (i.e., a multilevel or random effect model with fixed effects - including random effects); 5) weighted regression models, inluding mixed effect models; and 6) linear regression models (including linear and generalised linear mixed effect models, and phylogenetic least squares models). Notably, only model types 1-5 are weighted models.

@@

## **Opening Up New Opportunities in Comparative Physiology: Expanding the Breadth of Effect Sizes to Deal with "Nuisiance Heterogenity"**

@@

As is clear from Table \@ref(tab:tablea) none of the effect measures control for 'nusance heterogenity'. For example, the magnitude of response between two treatment groups for lnRR or SMD might depend on just how different the tempeartures, dosages, pH or water potential applied in each treatment. Even modelling just the mean often needs to account for 'nusance heterogenity'. For example, standard metabolic rate (SMR) is tempearture dependent. As such, one needs to standardise the temperature measurement to compare across studies [@White2006; @Uyeda2017, if, of course, temperature is not of interest in of itself]. While we could of course model this explictly, this may not always be desireable. It can often be useful to standardise the effect to both ease its interpretation and simplify modelling. In Box 1 we provide readers with guidance on how comparative physiologists can apply corrections to the effect measure itself. We highlight two examples in the box focusing mainly on generality. Below, we show using a common well understood effect measure of interest to comparative physiologys, $Q_{10}$ how we can formalise a common effect size measure $lnRR$ to capture this 'nusance variation'.

#### *Comparing Changes in Mean Physiological Rates, $Q_{10}$*
A very common occurrence in comparative physiology is to manipulate the temperature organisms experience and measure some physiological rate (e.g., metabolic rate). Using normal effect size measures (see above) would result in the effect sizes varying in the types of temperatures applied within and across across studies contributing to 'comparability' issues. As such, it is often of interest to standardize effects by temperature. One common effect size measure already used in comparative physiology to compare physiological rates is $Q_{10}$ [e.g., @Rodgers2021; @Seebacher2015; @Havird2020], which describes the multiplicative change in physiological rates across a 10°C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. Currently, however, we do not have associated sampling variances derived for $Q_{10}$ making it challenging to make use of the full power of weighted meta-analytic models. Interestingly, $Q_{10}$ can be seen as a variant of *lnRR*, meaning that we can derive a $Q_{10}$ based effect size and sampling error using the well known mathematical properties of *lnRR*. This opens up the meta-analytic toolbox and improves upon our ability to account for well known sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017].

@@

$$
\begin{equation} 
  s_{lnRR} = \left ( \frac{SD_{1}^2}{N_{1}X_{1}^2}\right ) + \left ( \frac{SD_{2}^2}{N_{2}X_{2}^2}\right )
  (\#eq:slnRR)
\end{equation} 
$$ In equation \@ref(eq:lnRR), $X_{1}$ is the mean of group 1, often a control group, where as $X_{2}$ is the mean of group 2 (e.g., a treatment group). The mean of $X_{i}$ for group *i* can be any measurement type (e.g. a physiological rate, mass etc) so long as the measurement variable is ratio scale. Log transformation of this ratio makes the effect size normally distributed. <!--- use consistent notations in relation to the table - do you want to put i or not - then you need to put i throughout ---> Equation \@ref(eq:slnRR) is the analytical solution for *lnRR*'s sampling variance where, $SD_{1}^2$ and $SD_{2}^2$ is the sample variance in group 1 and 2, respectively and $N_{1}$ and $N_{2}$ is the sample size in group 1 and 2.

@@

#### *Formalising Effect Size Metrics that Compare Changes in Variability Across Treatments in the Presence of "Nuisance Heterogenity"*

>>>>

A meta-analysis provides quantitatively informed answers to research questions by synthesising research results across studies and identifying sources of variation [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Arnqvist1995; @Koricheva2013; @Borenstein2019]. Meta-analysing independent studies provides greater statistical power and precision than what any individual study on its own would be able to provide on a question - particularly given that most empirical studies are already under-powered in many areas of biology [@Forstmeier2017; @Jennions2003; @Button2013]. By expressing study effects on a common scale (i.e., standardised effect size) a meta-analyst can gain broader insight into the direction and efficacy of a particular treatment effect or the strength of a relationship between two variables of interest [@Gurevitch2018; @Koricheva2013]. Quantitative syntheses, including meta-analyses, have provided comparative physiologists with powerful insights on pressing global challenges - from testing whether physiological plasticity can buffer organisms against climate change [e.g, @Seebacher2015] to the degree to which endocrine disrupting chemicals, such as bis-phenol A (BPA), impact aquatic organisms [e.g., @WuSeebacher2020].

@@

Despite its widespread adoption and well established methodological procedures, meta-analysis is often criticised for mixing 'apples and oranges' -- in other words, mixing effects from studies that lack comparability [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010; @Glass2015]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units, which can impact the generalisability of meta-analytic results [@Arnqvist1995; @Stewart2010]. Indeed, study specific differences in comparative physiology are widespread. For example, experimental designs can vary in the temperatures that they apply or in the dosages of chemical modifiers used in experimental treatments [e.g., @WuSeebacher2020]. To the uninitiated, this concern might not be unfounded given that heterogeneity in effects is often high in ecological and evolutionary meta-analyses [@Senior2016], telling us that the effect size varies a great deal from one population to another. However, to many, the very fact that highly heterogeneous effects exist is a core strength of meta-analysis [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015]. Indeed, exploring the reasons for why effects are likely to vary across studies "...is the spice of life" [p.g., 519; @Cooperetal2019].

The goal of our review is to briefly overview the process of meta-analysis, making a case for why heterogeneity in a meta-analysis is a good thing for answering fundamental research problems and guiding research directions. We introduce the idea of 'nuisance heterogeneity', showing comparative physiologists that it can be comfortably dealt with by re-formalising many existing effect sizes measures and/or by using multilevel meta-regression models. Indeed, a combination of these approaches might often be ideal. We first conduct a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic practices -- determining what types of effect sizes are commonly used, how often effect size estimates are impacted by 'nuisance heterogeneity' and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. We then formalise alternative effect sizes and their associated sampling variance to provide comparative physiologists opportunities to explicitly incorporate treatment differences into the effect size estimates to ease their interpretation. We then describe how more complex treatment differences, such as dosage differences (which may have non-linear effects) can be accommodated using multilevel meta-regression models. We hope to show how expanding the meta-analytic toolkit for comparative physiologists will provide new opportunities to address global scale problems that attempt to understand how organisms will cope with rapidly changing environments in the future.

@@

Generally, contributors to effect size heterogeneity fall within a number of broad categories. First, sampling variance contributes to heterogeneity; the variability driven by the inherent uncertainty in sampling a given population. We cannot sample the entire population and nor do we necessarily want to. As a consequence, we need to rely on a sub-sample to estimate a population parameter (e.g., correlation between two traits). How many independent samples we collect impacts our precision in estimating this 'true' effect [@Borenstein2019]. In meta-analysis we formally include sampling variance to remove this source of heterogeneity directly because it is known -- hence why having sampling variances for an effect is so important [@Gurevitch2018; @Koricheva2013]. Second, there are sociological factors impacting heterogeneity. These include attitudes towards publishing non-significant results. Low powered studies that do not find an 'effect' or that find an effect in the opposite direction often lead to what is called publication bias [@Jennions2013; @Nakagawa2021b; @Rothstein2005]. Third, there is real biology driving differences in effects. This is particularly true in comparative physiology where different species with varying habitats, life-histories, mating systems and reproductive modes are synthesised. As comparative physiologists, this is the stuff that gets us up every morning! Much variation could result from shared evolutionary history (i.e., phylogeny) or because of the different thermal environments species (or populations) occupy. We can model these biological factors to test predictions from hypotheses that we believe are key players impacting mean effect size and direction. This improves our understanding of the biological world around us and helps inform on future research directions. Finally, methodological factors are also big contributors to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it is important we know and understand this as it can determine how experiments are designed in the future. In other cases, methodological factors might be considered 'nuisance heterogeneity' [following from @Cooperetal2019] -- factors we know vary, but are rather unsurprising to us. These might include differences in temperature, pH, dosages, water potential etc. It might already be clear that these factors will have an impact on the magnitude of an effect, but they might distract from the core interest of a meta-analysis. In some cases, however, the latter factors might actually not be 'nuisance' variables but the main interest of a synthesis. However, in many cases they can result in a more complicated interpretation of an effect measure then what is desired.

All the various sources of heterogeneity (except sampling variance) all contribute (to varying extents) to the 'apples and oranges' problem often touted as being an issues with some meta-analyses (or quantitative syntheses more generally). Indeed, some of this criticism revolves around combining studies that vary in their quality (e.g., observational vs experimental). Combining standardised effects, while ignoring these sources of heterogeneity can result in the meaning of a meta-analysis not being particularly useful to many. However, many argue that this is just what we are interested in synthesizing [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015]. In the words of Gene Glass, the father of meta-analysis: "Of course it [meta-analysis] mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. @Glass2015's intention was to simply highlight that whether comparing apples and oranges is a good idea simply boils down to the specific question of interest, the population one wishes to make conclusions about and how heterogeneous effects actually are. If we are interested in gaining insight into 'fruit' then synthesising apples and oranges is the only thing that makes sense. Instead, if we are interested in generalising only to apples then mixing fruit will not be ideal. In addition, if we want to understand why different 'fruit' vary in their effects, then we can only ask this question by including all fruit [@Borenstein2019].

We tend to agree with @Glass2015, and his sentiment is of particular relevance to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms then including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are known to differ. By synthesizing diverse study outcomes, it allows us to not just report on overall effects (if the question is relevant to ask), but to understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction) and approaches that can explain the diversity of effect outcomes observed. In doing so, it provides a richer set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included within a meta-analysis, and how this can affect interpretations made -- an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. If the estimate of an overall effect coincides with low heterogeneity estimates then the effects are reasonably consistent [@Borenstein2019]. If, however, an overall effect is accompanied by large heterogeneity estimates, then we would shift focus and explore drivers of that heterogeneity. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant, and indeed may be misleading. Therefore, understanding overall effects in each of these groups separately (and whether they differ) will be more useful. It is important to consider these questions when conducting meta-analyses.

Of course, there are certain types of heterogeneity that transcend different questions or that are simply a 'nuisance' to interpreting effects. In other words, if we compare different types of 'apples' is it still fair to synthesize these studies to understand overall effects? This situation might be analogous to a meta-analysis in comparative physiology that attempts to understand the effects of diet on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of fish under different temperatures, but in the end, they are still all the same fish species, governed by the same thermodynamic effects on growth. In this situation, we might still be interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. It is not always clear how we should tackle these sources of heterogeneity. If temperature differences are of direct interest, then we should definitely model them. If, however, it is less interesting, than say the diet administered, then it may introduce an unnecessary complication to the interpretation of the effect(s) of interest. Here, we focus on meta-analytic tools to circumvent these issues and demonstrate that the 'apples and oranges' problem is not really a problem at all.

@@

Before embarking on "meta-analytic enlightenment" it is worth first understanding how comparative physiologists are currently conducting meta-analyses / comparative analysis.  As discussed above, different types of questions and experimental approaches will mix 'apples and oranges' to varying extents, and indeed, this heterogeneity (or basket of fruit) is often the main interest to comparative biologists. To better understand the different types of effect sizes being used, their susceptibility to known sources of methodological / environmental variability across studies, and how these sources are controlled we conducted a literature survey. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to continuous variation in how treatments were applied (e.g., for contrast-based effects) or the conditions under which traits were measured (i.e., for correlations) across studies (i.e., 'nuisance heterogeneity')? 3) If so, was the variation accounted for in the effect size or during analysis?, and finally, 4) how were effect sizes analysed in the quantitative synthesis, were these weighted analyses, and what program was used for analysis? We collected additional

@@

All searches were conducted in Scopus between May 16 - 18, 2021. Briefly, we first did a series of pilot searches (detailed in Supplementary Materials) to refine our search query [See @Foo2021 for an overview of this process]. We searched for studies that included: 'meta-analysis', 'meta regression' , 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macro physiology' in their title and abstract, restricting our search to key comparative physiology journals (detailed in Table S1 and S2 - Supplementary Materials) within the last 6 years. We expected this to provide a contemporary overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We acknowledge that this may have missed some important, and influential, quantitative synthesis in the field of comparative physiology, but likely provided a random sample to inform on the questions we sought to address (see below).

We included quantitative syntheses, including meta-analyses (*sensu stricto*), that were focused on comparative physiological questions only. Narrative reviews and comments were not considered because they did not collate quantitative data from the literature. Studies needed to have collected physiological trait data from the literature including traits, such as hormones, thermal physiology (CTmax, CTmin, thermal preferences, thermal breadth, LT50, lethal thresholds), metabolic rates (routine, maximum, and minimum metabolic rate, aerobic scope), enzyme reaction rates and concentrations, oxidative stress and consequences (e.g., DNA damage, teleomere attrition), performance measurements (e.g., heart rates, sprint speed, swim speed, growth rates), body composition (e.g., fatty acid profiles) and immune traits (e.g., cell function, antibodies, ect.). Studies were also included if they measured the relationship between physiological traits and body size (e.g., metabolic scaling), reproductive allocation, or survival / mortality across taxa. Studies on behavior were only included if the synthesis was explicitly interested in the proximate physiological drivers of behavioral variation. We also included meta-analyses that quantified plasticity in physiological traits across taxa for the purpose of understanding adaptive plasticity in response to environmental variation. While many studies in comparative physiology collect data as part of a comparative analysis, we focus here only on studies collating data from existing literature because this is the major data source for quantitative synthesis. Quantitative syntheses interested in phenology, bio-mechanics, sexual size dimorphism (i.e., SSD, except if the study was interested in sexual differences in physiological traits), those focused on validating physiological methodology, or characterising population genetic structure, gene transcription and quantitative genetics were excluded. There were also meta-analyses that focused on colour evolution and contrast which varied in their goals but were most often not focused on physiological questions. As such, these studies were also excluded. We restricted our focus to animal studies because plant synthesis papers were often only focused on ecological (e.g., quantified changes in abundance and distribution of species) or agricultural questions (e.g., studies on soil microbial function, nitrogen and phosphorus capture, methane output and greenhouse gas emissions).

All title, abstract and full text screening was done by two people (DWAN and ML). Any conflicts or disagreements were resolved by discussion and a consensus on a paper was reached. Out of the 417 papers, 79 full texts were screened and we included a total of 62 for final data extraction. For full details on the search strings used, dates searched, the number of papers found, and our PRISMA flowchart (along with reasons for exclusion at the full text stage) we refer readers to the *Supplementary Materials*. To address our questions, we categorised effect sizes as falling into the following categories: Response Ratio (lnRR), Standardised Mean Differences (SMD; e.g., Hedges d, g or Cohen's d), Correlation or Fisher's Z -transformed Correlation, Regression Slopes (Slopes), Risk Ratios, or whether the paper analysed the raw data or summary statistics directly (i.e., means/proportions) (which we label as raw/mean/proportion). For all other effects we labelled as 'other'. To determine whether effect sizes were likely to vary based on some underlying continuous variable or were dependent on the context in which traits were measured, we assessed whether authors: 1) acknowledged this explicitly in the paper or 2) accounted for it in their models. If it was not clear, two of us (DWAN & PP) discussed the paper, and made a decision. For example, trait means, such as ectothermic metabolic rate (i.e., raw/mean/proportion-based effect size), will vary depending on the temperature being measured, the magnitude of differences between two treatments (e.g, contrast-based effect sizes - SMD or lnRR) might depend on the temperature or dosage used in treatments, and the strength of relationship between two variables (e.g., correlation between behaviour and metabolism) might depend on the temperature context under which metabolic rate or behaviour were measured. Given the high heterogeneity in effects used in comparative physiology, we expected this to be common to most effect sizes, however, it has not been formally quantified. We categorised the types of statistical models fit in each study as: 1) fixed-effect meta-analysis; 2) random effects meta-analysis (i.e., intercept only model, but has random study effects); 3) multi-level meta-analysis (i.e., intercept only model, but has multiple random effect levels such as study, species, observation-level random effects); 4) meta-regression models (i.e., a multilevel or random effect model with fixed effects - including random effects); 5) weighted regression models, including mixed effect models; and 6) linear regression models (including linear and generalised linear mixed effect models, and phylogenetic least squares models). Notably, only model types 1-5 are weighted models.

@@

## **Opening Up New Opportunities in Comparative Physiology: Expanding the Breadth of Effect Sizes to Deal with "Nuisance Heterogeneity"**

@@

As is clear from Table \@ref(tab:tablea) none of the effect measures control for 'nuisance heterogeneity'. For example, the magnitude of response between two treatment groups for lnRR or SMD might depend on just how different the temperatures, dosages, pH or water potential applied in each treatment. Even modelling just the mean often needs to control for various sources of 'nuisance heterogeneity'. For example, standard metabolic rate (SMR) is temperature dependent. As such, one needs to standardise the temperature measurement to compare across studies [@White2006; @Uyeda2017] (if, of course, temperature is not of interest in of itself). While we could of course model this explicitly, this may not always be desirable. It can often be useful to standardise the effect to both ease its interpretation and simplify modelling. In Box 1 we provide readers with guidance on how comparative physiologists can apply corrections to the effect measure itself. We highlight two examples in the box focusing mainly on generality. Below, we show using a common, well understood effect measure of interest to comparative physiology, $Q_{10}$ how we can formalise a traditional effect sizes', such as $lnRR$, to capture this 'nuisance variation'.

#### *Comparing Changes in Mean Physiological Rates, $Q_{10}$*
A very common occurrence in comparative physiology is to manipulate the temperature organisms experience and measure some physiological rate (e.g., metabolic rate). Using normal effect size measures (see Table \@ref(tab:tablea)) would result in the effect sizes varying in the types of temperatures applied within and across across studies contributing to 'comparability' issues. As such, it is often of interest to standardize effects by temperature. One common effect size measure already used in comparative physiology to compare physiological rates is $Q_{10}$ [e.g., @Rodgers2021; @Seebacher2015; @Havird2020], which describes the multiplicative change in physiological rates across a 10°C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. Currently, however, we do not have associated sampling variances derived for $Q_{10}$ making it challenging to make use of the full power of weighted meta-analytic models. Interestingly, $Q_{10}$ can be seen as a variant of *lnRR*, meaning that we can derive a $Q_{10}$ based effect size and sampling error using the well known mathematical properties of *lnRR*. This opens up the meta-analytic toolbox and improves upon our ability to account for well known sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017].

@@

$$
\begin{equation} 
  s_{lnRR} = \left ( \frac{SD_{1}^2}{N_{1}X_{1}^2}\right ) + \left ( \frac{SD_{2}^2}{N_{2}X_{2}^2}\right )
  (\#eq:slnRR)
\end{equation} 
$$ In equation \@ref(eq:lnRR), $X_{1}$ is the mean of group 1, often a control group, where as $X_{2}$ is the mean of group 2 (e.g., a treatment group). The mean of $X_{i}$ for group *i* can be any measurement type (e.g. a physiological rate, mass etc) so long as the measurement variable is ratio scale. Log transformation of this ratio makes the effect size normally distributed (most commonly assumed by meta-analytic models). Equation \@ref(eq:slnRR) is the analytical solution for *lnRR*'s sampling variance where, $SD_{1}^2$ and $SD_{2}^2$ is the sample variance in group 1 and 2, respectively and $N_{1}$ and $N_{2}$ is the sample size in group 1 and 2.

@@

#### *Formalising Effect Size Metrics that Compare Changes in Variability Across Treatments in the Presence of "Nuisance Heterogeneity"*
